### 2016-03-02 CICM2016/appendices.tex
Updates

\iffalse
%\iftrue
\section*{Appendix: Syntax and Translation}

\begin{figure}
  \begin{equation}
    \begin{split}
      expr\    \rightarrow\ & \CVar\ id                          \\
                         |\ & \CLit\ literal                     \\
                         |\ & \CApp\ expr\ expr                  \\
                         |\ & \CLam\ \mathcal{L}\ expr           \\
                         |\ & \CLet\ bind\ expr                  \\
                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                         |\ & \CType                             \\
      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                         |\ & \CGlobal\      \mathcal{G}         \\
                         |\ & \CConstructor\ \mathcal{D}         \\
      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                         |\ & \CLitStr\ \mathcal{S}              \\
      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                         |\ & \CLitAlt\ literal                  \\
                         |\ & \CDefault                          \\
      bind\    \rightarrow\ & \CNonRec\ binder                   \\
                         |\ & \CRec\ [binder]                    \\
      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
    \end{split}
  \end{equation}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers \\
    $\mathcal{D}$ & constructor identifiers
  \end{tabular}

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
  \label{fig:coresyntax}
\end{figure}

\begin{figure}
  \begin{align*}
    toTree(e) &=
    \begin{cases}
      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
      (\feature{\CType})                                                & \text{if $e = \CType$} \\
      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
    \end{cases}
  \end{align*}
  \caption{Transforming Core expressions of Figure \ref{fig:coresyntax} to rose trees. The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support. We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.}
  \label{fig:totree}
We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
\end{figure}

>>>>

%\iffalse  % WITHOUT APPENDICES
\iftrue    % WITH APPENDICES

\appendix

\section{Core Syntax}\label{sec:core}

We use a simplified syntax for the GHC Core language, given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively). We mechanically translate the full GHC Core language into this sub-set, which does not contain types, prior to performing recurrent clustering.

%\begin{figure}
  \begin{equation}\label{fig:coresyntax}
    \begin{split}
      expr\    \rightarrow\ & \CVar\ id                          \\
                         |\ & \CLit\ literal                     \\
                         |\ & \CApp\ expr\ expr                  \\
                         |\ & \CLam\ \mathcal{L}\ expr           \\
                         |\ & \CLet\ bind\ expr                  \\
                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                         |\ & \CType                             \\
      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                         |\ & \CGlobal\      \mathcal{G}         \\
                         |\ & \CConstructor\ \mathcal{D}         \\
      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                         |\ & \CLitStr\ \mathcal{S}              \\
      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                         |\ & \CLitAlt\ literal                  \\
                         |\ & \CDefault                          \\
      bind\    \rightarrow\ & \CNonRec\ binder                   \\
                         |\ & \CRec\ [binder]                    \\
      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
    \end{split}
  \end{equation}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers \\
    $\mathcal{D}$ & constructor identifiers
  \end{tabular}
%\end{figure}

\section{Rose Trees}\label{sec:rosetree}

The $toTree$ function shown below transforms Core expressions, described in Appendix \ref{sec:core} to rose trees. We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.

The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support. We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.

%\begin{figure}
  \begin{align*}\label{fig:totree}
    toTree(e) &=
    \begin{cases}
      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
      (\feature{\CType})                                                & \text{if $e = \CType$} \\
      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
    \end{cases}
  \end{align*}

%\end{figure}

### 2016-03-02 CICM2016/background.tex
Updates

plus :: Nat -> Nat -> Nat
plus    Z  y = y
plus (S x) y = S (plus x y)

mult :: Nat -> Nat -> Nat
mult    Z  y = Z
mult (S x) y = plus y (mult x y)

@@

\begin{figure}
  \begin{small}
    \underline{\texttt{plus}}
    \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Local "y")))
                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
                                               (App (App (Var (Global "plus"))
                                                         (Var (Local  "x")))
                                                    (Var (Local "y"))))
                                          "x")))
    \end{verbatim}

    \underline{\texttt{mult}}
    \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                    (Var (Local  "y")))
                                               (App (App (Var (Global "mult"))
                                                         (Var (Local  "x")))
                                                    (Var (Local  "y"))))
                                          "x")))
    \end{verbatim}

    \underline{\texttt{odd}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}

@@

\begin{align*}
  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
\end{align*}

@@

\begin{align*}
  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
\end{align*}

Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments. In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.

\begin{figure}
  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
  \begin{verbatim}
                  plus a b = plus b a
                  plus a Z = a
         plus a (plus b c) = plus b (plus a c)
                  mult a b = mult b a
                  mult a Z = Z
         mult a (mult b c) = mult b (mult a c)
              plus a (S b) = S (plus a b)
              mult a (S b) = plus a (mult a b)
         mult a (plus b b) = mult b (plus a a)
                 odd (S a) = even a
            odd (plus a a) = odd Z
           odd (times a a) = odd a
                even (S a) = odd a
           even (plus a a) = even Z
          even (times a a) = even a
plus (mult a b) (mult a c) = mult a (plus b c)
  \end{verbatim}
  \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
  \label{fig:qspecresult}
\end{figure}

These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}. As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
\fi

>>>>

@@

\begin{figure}
  \begin{small}
    \underline{\texttt{odd}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}

@@

\begin{align*}
  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{odd}, \texttt{even}\}
\end{align*}

@@

\begin{align*}
  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
                     & \texttt{S}\ c,\ \texttt{odd Z},\ \texttt{odd}\ a,\ \dots ]
\end{align*}

Notice that functions such as \hs{odd} and \hs{even} are valid terms, despite not being applied to any arguments. In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.

These terms will be grouped into four classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}. As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations such as $odd (S a) = even a$ and $even (S a) = odd a$ are conjectured.
\fi

### 2016-03-02 CICM2016/contributions.tex
Updates

\begin{figure}
    \begin{equation*}
      (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
     \end{equation*}
     \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
     \label{fig:vectorexample}
\end{figure}

These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, with any required padding added on the right. These matrices are either padded with $0$ or truncated, to fit a fixed size $r \times c$.

\begin{sloppypar}
Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_1 \cat \mathbf{M}_2 \cat \dots \cat \mathbf{M}_r$, as shown in Figure \ref{fig:vectorexample}:
\end{sloppypar}

\begin{equation}
  featureVec(e) = pad(level(1, toTree(e))) \cat \dots \cat pad(level(r, toTree(e)))
\end{equation}

@@

To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which gives equal values for $\alpha$-equivalent expressions (i.e. renaming an identifier shouldn't affect the feature). To do this, the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from Figure \ref{fig:totree} for clarity). The context is a list of the local identifiers which are in scope, prepended as they are encountered. The context is initially empty, and only extended when $toTree$ calls itself recursively.

For example, when calculating $toTree(\CLam\ i\ e)$ in context $\vect{x}$, we make the recursive call $toTree(e)$ in the context of \emph{$\vect{x}$ prepended with $i$} (i.e. $(i) \cat \vect{x}$). Similar extensions of the context are performed in the cases of \CBind, \CCase\footnote{The $\mathcal{L}$ value in a \CCase\ expression is bound to the expression being matched against; a technical detail to preserve sharing.}, \CAlt\ (which may introduce an arbitrary number of local identifiers) and \CLet\ (where identifiers are taken from occurrences of \CBind\ in the first expression).

Well-formed Haskell declarations do not contain free variables (or, equivalently, free variables are encoded as global identifiers rather than local identifiers). Hence if we apply $toTree$ to (the Core expression corresponding to) such declarations we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$. In which case we define:

\begin{equation} \label{eq:localfeature}
  \feature{l} = i + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
\end{equation}

Where $i$ is the index of the first occurrence of $l$ in the context. This tells us how many binders a variable is nested inside, relative to the point it was introduced. This is invariant under renaming, as we wanted, and is known as the \emph{de Bruijn index} of $l$. We again use $\alpha$ to separate these features from those of other constructs.

Since the $toTree$ function discards numerals, strings and constructor identifiers, the only remaining case is global identifiers $\mathcal{G}$. Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers. We also cannot directly encode the form of the identifiers, e.g. using a scheme like G{\"o}del numbering, since this is essentially arbitrary and has no effect on their semantic meaning (referencing other expressions).

@@

Where $\vect{C}$ are our clusters (in some arbitrary order). This is where the recurrent nature of the algorithm appears: to determine $\vect{C}$ we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.

For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference). In this way, we can avoid looking up expressions which haven't been clustered yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations. If an identifier cannot be found in any cluster, it must appear in the same SCC as the expression we are processing; hence we use the constant feature value $f_{recursion}$.

>>>>

These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, read left-to-right, followed by any required padding. These matrices are then either truncated, or padded with rows (on the bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.

\begin{sloppypar}
Finally, matrices are turned into vectors by simply concatenating the rows from top to bottom, hence the matrix from Figure \ref{fig:matrixexample} will begin $(\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, \dots$.
\end{sloppypar}

@@

To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which gives equal values for $\alpha$-equivalent expressions (i.e. renaming an identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn index} of the identifier \cite{de1972lambda}, denoted $i_l$:

\begin{equation} \label{eq:localfeature}
  \feature{l} = i_l + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
\end{equation}

We again use $\alpha$ to separate these features from those of other constructs.

Since numerals, strings and constructor identifiers are discarded during conversion to rose trees, the only remaining case is global identifiers $\mathcal{G}$. Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers. We also cannot directly encode the form of the identifiers, e.g. using a scheme like G{\"o}del numbering, since this is essentially arbitrary and has no effect on their semantic meaning (referencing other expressions).

@@

Where $C_i$ are our clusters (in some arbitrary order). This is where the recurrent nature of the algorithm appears: to determine the contents of $C_i$, we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.

For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference). In this way, we can avoid looking up expressions which haven't been clustered yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations. If an identifier cannot be found amongst those clustered so far, it must appear in the same SCC as the expression we are processing; hence we give that identifier the constant feature value $f_{recursion}$.

### 2016-03-02 CICM2016/future.tex
Updates

Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made. Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further. In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique. Below, we elaborate on the details, background and motivation for these choices.

\subsection{Clustering Extensions}
\label{sec:preprocessing}

The most glaring omission in our algorithm is its disregard for types. By ignoring types, not only are we losing valuable information about expressions, but we also lose the ability to distinguish between constructors. This is because a constructor, like \hs{True} or \hs{Just}, has no internal structure; it is just a token. The distinguishing features of constructors are their types, which not only tell us which data type they construct, but also their arity, the types of their arguments, etc.

Our algorithm closely follows that of ML4PG, which \emph{does} support types. This is handled by populating matrix cells with tokens \emph{and} their types. Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).

One partial solution would be leave most Core expressions without types, but to include them for non-local identifiers (i.e. globals and constructors), which we can look up in a database. In fact, our ML4HS framework already includes such type information in its database, alongside the Core syntax trees. Integrating this information into our algorithm is the next logical step.

@@

With more expressive features, it may also be useful to experiment with more powerful learning algorithms. An interesting possibility is to add a feedback loop between the theory exploration phase and the clustering phase, to more directly base the similarity of expressions on whether they (are predicted to) occur together in equations.

\subsection{Theory Exploration Extensions}

Our current approach is a rather conservative change to the existing theory exploration approaches, as it is essentially a wrapper around \qspec{}. There is potential for more radical changes to be made, which alter the search process itself.

\subsubsection{Variable Instantiation}

\qcheck{} is certainly the most popular property checker for Haskell, which motivates its use in \qspec{} to instantiate variables to random values. However, this task of finding type inhabitants has also been solved in many other ways, which may be worth investigating in place of \qcheck{} (or perhaps even as part of an ensemble).

The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly. Whilst this does not make \textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage is that it may use much less memory, as the generated values are built up incrementally. In contrast, \qcheck{} may generate very large values; in particular, generating tree structures na\"{\i}vely can cause them to grow exponentially. For example, here is a potential generator for \hs{RoseTree}s:

\begin{haskell}
genRoseTree = do f        <- arbitrary
                 subtrees <- listOf genRoseTree
                 return (Node f subtrees)
\end{haskell}

The \hs{listOf genRoseTree} call will return a list of arbitrary length, where each element is generated by \hs{genRoseTree}. This allows an arbitrary number of recursive calls to be made for each invocation of \hs{genRoseTree}, which will quickly exhaust the resources of any machine. Whilst such problems may be anticipated, or quickly spotted, in a property checking setting, this can be more difficult for our automated approach. For example, if a type does not have a generator available, we cannot use a library like \hs{derive} to create one automatically, as it suffers from this na\"{\i}vity problem.

A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \cite{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded. This may narrow down our search procedures greatly, especially when predicates are involved. \qcheck{} allows predicates to restrict the values it tests with, and hence allows \emph{conditional} equations to be discovered. However, its implementation uses a simple rejection sampling technique: values are generated just as if the predicate were not there, and afterwards are filtered to reject any which do not satisfy the predicate. This makes it difficult to use very specific predicates, as it is unlikely that many of our random samples will exactly match our criteria. On the other hand, \textsc{Lazy SmallCheck} will focus its search on exactly those parts of the datastructure which are checked by the predicate, as those are the parts being forced to evaluate. This makes it much more likely that we will find values which satisfy the predicate, allowing us to effectively explore more specific conditional properties.

Other approaches to generating inhabitants include \textsc{Djinn} \cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator). \textsc{MuCheck} \cite{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g. changing the order of pattern match clauses); whilst not as exhaustive as the other approaches, mutating existing values in this way is claimed to yield values which correspond more closely to what a programmer might write. This is an interesting possibility for focusing theory exploration on to more ``realistic'' areas of the search space, and hence avoiding some of the more useless or bizarre expressions that random search and enumeration may produce.

In fact, the database generated by our \textsc{AstPlugin} may prove helpful in generating values, since its type information can be fed to a tool like \textsc{Djinn} to discover chains of function applications for building values, which would be particularly useful in cases where constructors are private, like in our email example. This is similar to the \textsc{Hoogle} tool, but also offers the ability to use dependency information to avoid potentially infinite recursion.

The Core syntax trees in our database could also be used to generate theories for automated theorem provers. \hspec{} currently uses the GHC API to transform Core within its own process, however that approach suffers from the problems described in \S \ref{sec:astplugin}.

\subsubsection{Interestingness}
\label{sec:interestingness}

If we do succeed in producing a fast theory exploration system, which chooses productive combinations of terms and finds a large number of properties, we encounter the problem of managing the output: finding the needles we are interested in among the haystack of trivialities and coincidences.

This is governed by the ``interestingness'' criteria of the theory exploration system: what to keep and what to discard, and even what areas of the search space to prioritise. \qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is very simple: we discard equations which are direct consequences of others, and keep all the rest. Different, and more sophisticated notions of interestingness have been widely studied in other fields, which may be applied in the context of theory exploration.

>>>>

Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made. Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further.

The most obvious next step for our algorithm is to incorporate types. Types contain valuable information about an expression, and allow us to distinguish between constructors. Our algorithm closely follows that of ML4PG, which \emph{does} support types, by populating matrix cells with tokens \emph{and} their types. Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).

@@
